---
title: "Day 2"
author: "Ellen"
date: "12/1/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Tidyr 
Making long data wide, and wide data long 
```{r, include=FALSE}
library(tidyverse)
```

##Example Datasets
```{r example}
gap_wide <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")


##yesterdays gapminder data
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

head(gap_wide)
head(gapminder)

```
##gather()

Use gather to turn 'gap_wide' into a long format dataset
Key and value. One column is the key and the other will be the value
:: - telling R to use a package for a single function

```{r gather}

gap_long <- gap_wide %>% 
  gather(key=obstype_year, value=obs_values)

head(gap_long)
tail(gap_long)

##this is not what we want to do. Because want to keep those first two columns (continent and countr) and want other columns next to that 

gap_long <- gap_wide %>% 
  gather(key=obstype_year, value=obs_values, 
         dplyr:: starts_with("pop"), 
         dplyr::starts_with("lifeExp"), 
         dplyr::starts_with("gdpPercap"))

head(gap_long)
```

###Keep columns with '-' in gather()
```{r better way to do it}

gap_long <- gap_wide %>% 
  gather(key=obstype_year,
         value=obs_values, 
  -continent, -country)
head(gap_long)

```

###separate out the year
```{r}
gap_long <- gap_wide %>% 
  gather(key=obstype_year,
         value=obs_values, 
  -continent, -country) %>% 
  separate(obstype_year,
           into=c("obs_type", "year"), sep="_",
           convert=T)

##convert=T turns year from a character into an integer

head(gap_long)
```

##Plot long format data
```{r plotting}
life_df <- gap_long %>% 
  filter(obs_type=="lifeExp",
         continent=="Americas")

head(life_df)

ggplot(data=life_df, aes(x=year, y=obs_values, color=country)) + geom_line()

```

Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: use the dplyr::group_by() and dplyr::summarize() functions.

```{r}

life_df <- gap_long %>% 
  filter(obs_type=="lifeExp") %>% 
  filter(year %in% c(1982:2007))

head(life_df)
tail(life_df)

group_example=life_df %>% 
  dplyr:: group_by(continent, year) %>% 
  dplyr:: summarize(mean=mean(obs_values)) 
  

ggplot(data=group_example, aes(x=year, y=mean, color=continent)) + geom_line() + theme_bw()
```

##spread()

Getting the dataset back into the wide format to look like original gapminder 
```{r}

gap_normal <-  gap_long %>% 
  spread(key=obs_type, value=obs_values)

head(gap_normal)


```

